Work Fragmentation in Developer Interaction Data

### 2015-12-16 paper_2.tex
New structure, first version

\begin{table}[ht!]
\tiny
\renewcommand{\arraystretch}{1.3}
\caption{Kinds of interaction events~\cite{KM06}. }
\label{tbl:kind_event}
\centering
\begin{tabular}{|p{1.7cm}|p{1.3cm}|p{5cm}|} 
  \hline 
kind & mode & description \\  
  \hline 
    \hline 
selection & direct &  Editor and view selections via mouse \\
edit & direct & Textual and graphical edits  \\
command & direct & Operations such as saving, building, preference setting  \\
manipulation & direct & Direct manipulation of interest  \\
propagation & indirect & Interaction propagates to structurally related elements  \\
prediction & indirect & Capture of potential future interaction events \\
  \hline
\end{tabular}

@@

\begin{table}[ht!]
\tiny
\renewcommand{\arraystretch}{1.3}
\caption{ Example of interaction history. }
\label{tbl:sample_event}
\centering
\begin{tabular}{|c|l|c|c|c|} 
  \hline 
    & StartTime & EndTime & EventKind &  Method  \\
  \hline
  \hline
  1 & 10:30:00 & - & selection & m1  \\
  2 & 10:30:40 & - & manipulation & - \\
  3 & 10:30:40 & - & edit & m1  \\
  4 & 10:31:03.700 & - & selection & m1  \\
  5 & 10:31:03.800 & - & selection & m2  \\	  
  6 & 10:31:03.850 & - & selection & m3  \\	  
  7 & 10:32:05 & 10:33:07 & edit (5) & m2  \\	  
  8 & 10:33:10 & - & prediction & - \\
  \hline  
\end{tabular}
\end{table}

@@

\section{Preview Phases for Knowledge Discovery}
We used the phases of the process \textit{Knowledge Discovery in Databases (KDD~\cite{FPP96})} to discover useful knowledge from our collection of data: we first select data, then preprocess it, before transforming it to time series. We close this section by presenting the metrics we use in this study. %  (see Figure~\ref{fig:esq}).

\subsection{Selection}
Our purpose is the analysis of work fragmentation and interruptions in software development. Therefore, we focused only in the interaction data of the user with the program elements. The  variables of interest for our study are~\cite{RD13}:
\begin{enumerate}
\item Duration: the time that a programmer took to perform a programming task. 
\item Edition:  the amount of code changes that were necessary to perform said task. %It is associated to the  edit events.
\item Navigation: how many program elements were consulted in a task; represented by selection events.
\item Edit ratio: as in previous studies \cite{KM06}, the ratio of edits over edits and selections is an indicator of more efficient work since program exploration is reduced.
\end{enumerate}

@@

 \begin{table}[hb!]
 \tiny
 \renewcommand{\arraystretch}{1.3}
 \caption{Number of events after preprocessing.}
 \label{tbl:mass_events}
 \centering
 \begin{tabular}{l | c | c | c } 
   type event & before & after & \% variation  \\  
   \hline 
   edit &	 452236 & 475554 &	+5\%   \\
   selection &	658768 & 527652 & -20\%   \\
   \hline
   total & 1111004 & 1003206 & -10\%  \\
 \end{tabular}
 \end{table}

@@

\begin{table}[ht!]
\tiny
\renewcommand{\arraystretch}{1.3}
\caption{Prevalence of interruptions according to their duration time. }
\label{tbl:by_duration}
\centering
\begin{tabular}{p{3cm}|p{0.8cm}|p{4cm}} 
duration & ratio & examples  \\
  \hline   
$[3  - 12 min \rangle$ & 69\% & short pause, answer a question, thinking, looking for interruption  \\ 
\hline 
$[12  - 30 min \rangle$ & 18\% & coffee break, short meeting, extended interruption  \\
$[30 min - 2 hr \rangle$  & 9\% & lunch break, a meeting  \\
$[2 hr - 8 hr \rangle$ & 4\% & extended meeting \\
%$[8 hr - \infty \rangle$ & 0.14 & \textit{a long interruption}   \\
%$[8 hr -  \infty \rangle$ & 0.06  &  a weekend \\
%$[2 day - \infty \rangle$ & 0.04 & a long holiday \\ 
\end{tabular}
\end{table}

\subsection{Metrics Used in This Study}
We use the following six metrics to measure interruptions and productivity, while controlling for the unproductive time in a session, the length of time of the session itself, and the efficiency of the development that took place during the session:

@@

\begin{figure}[t]
\centering
\begin{tabular}{c}
\includegraphics[width=.5\linewidth,clip=, angle=0]{figures/trans} 
\end{tabular}
\caption{Example of how to compress a time series. }
\label{fig:trans}
\end{figure}

\section{RQ1: Relationship between Interruptions and Productivity}
%\subsection{Effect in the productive work time}

@@

%First, we analyzed the number of interruptions over the productivity. In this sense, 
We split the data in five groups: The first group contain all the sessions without interruptions (\textit{none}). For the others groups, we have considered four ranges of number of interruptions delimited by their quartiles (Table \ref{tbl:quartil_int}). Then, for each group, we display the distribution of events per minute and edit ratio via boxplots (Figure \ref{fig:box_int_events}). We observe a large difference between the sessions without interruptions and the ones who do. Further, we observe that the rate of events per minute decreases slightly when the session has more interruptions. Therefore, we can intuit that the relationship between number of interruptions and the productivity indicators that are edits and selections, tends to be inversely proportional.

\begin{table}[ht!]
\tiny
\renewcommand{\arraystretch}{1.3}
\caption{Thresholds used to group sessions based on their number of interruptions} %separate the data in four percentages groups regarding the number of interruptions in each trace.}
\label{tbl:quartil_int}
\centering
\begin{tabular}{l | p{0.6cm} | p{0.6cm} | p{0.6cm}} 
     & $25\%$ & $50\%$ & $75\%$ \\  
  \hline 
  quartile &  3 & 5 & 10  \\   
\end{tabular}
\end{table}

\begin{figure}[!ht]
\centering
\begin{tabular}{c}
\minipage{0.33\textwidth}
\includegraphics[width=\linewidth,clip=, angle=0]{figures/box_int_events_edit}  
\endminipage\hfill
\minipage{0.33\textwidth}
\includegraphics[width=\linewidth,clip=, angle=0]{figures/box_int_events_sel}  
\endminipage\hfill
\minipage{0.33\textwidth}
\includegraphics[width=\linewidth,clip=, angle=0]{figures/box_int_ratio} 
\endminipage\hfill
\end{tabular}
\caption{Boxplots showing the relation between the number of edits and selection events per minute, the edit ratio, and the number of interruptions. }
\label{fig:box_int_events}
\end{figure}

Beyond visual inspection, we also quantify the statistical and the practical significance of these observations. First, all the differences observed are statistically significant with very low p-values (see Table~\ref{tbl:p_value}) according to the Mann-Whitney U-test. This is not surprising, given the shape of the boxplots and the size of the samples.

@@

@@

\begin{table}[ht!]
\tiny
\renewcommand{\arraystretch}{1.3}
\caption{Effect size and significance of the relationship between number of edit per minute, selection per minute, edit ratio, and number of interruptions} 
\label{tbl:p_value}
\centering
\begin{tabular}{l | p{0.75cm} | p{1.2cm} | p{1.2cm} | p{1.2cm} |p{1.2cm}}

@@

We first analyze the relationship between edit ratio and number of interruptions (Figure \ref{fig:box_int_events}). We observe that the edit ratio decreases when the session has more interruptions: the effect is pronounced between session that do not have interruptions and ones that do, and is more subtle as the number of interruption grows. A look at the practical and statistical significance of these results (Table \ref{tbl:p_value}, bottom) show that the results are (unsurprisingly) statistically significant, and that the observed effect sizes are large to very large.

In this section we analyze whether or not the interruption duration is a factor in the relationship between interruptions and developer productivity. To substantiate this claim, we have built two groups of sessions with interruptions:
\begin{itemize}
\item \textit{short}: the first group consists of sessions that only have short interruptions ($<$ 12 minutes of duration). These sessions constitute 18\% of the total.
\item \textit{prolonged}: the second group consists of the remaining sessions, which have at least one prolonged interruption ($\geq$ 12 minutes of duration). These session constitute 80\% of the total.
\end{itemize}

We then displayed the distribution of events by minutes and edit ratio of these two groups and compared them with the sessions without interruptions (Figure \ref{fig:box_dur_events}). We observe that the number of events per minute is lower in sessions with at least one prolonged interruption.

As in the previous case, the differences are significant, and we used \textit{Cohen's d} to measure the practical significance of the means (Table \ref{tbl:p_value2}). We note that the effect size of the interruption duration over the number of edits per minute is very large. In selections, the effect is moderate for short interruptions, and large for prolonged interruptions. We conclude that the relationship between user activity and interruptions could be adversely affected by interruptions of longer duration.

\begin{figure}[!ht]
\centering
\begin{tabular}{c}
\minipage{0.33\textwidth}
\includegraphics[width=\linewidth,clip=, angle=0]{figures/box_plot_range_edit}  
\endminipage\hfill
\minipage{0.33\textwidth}
\includegraphics[width=\linewidth,clip=, angle=0]{figures/box_plot_range_sel}  
\endminipage\hfill
\minipage{0.33\textwidth}
\includegraphics[width=\linewidth,clip=, angle=0]{figures/box_dur_ratio}
\endminipage\hfill
\end{tabular}
\caption{Boxplots showing the relation between the number of edits and selections per minute, the edit ratio and interruption duration}
\label{fig:box_dur_events}
\end{figure}

%\begin{figure}[!ht]
%\centering
%\begin{tabular}{c}
%\includegraphics[width=.8\linewidth,clip=, angle=0]{figures/box_int_ratio} \\
%\end{tabular}
%\caption{Boxplots of traces for showing the ratio of editions compared to the number of interruptions (top) and to the interruption duration (bottom). }
%\label{fig:box_ratio}
%\end{figure}

\begin{table}[ht!]
\tiny
\renewcommand{\arraystretch}{1.3}
\caption{Effect size and statistical significance of the relationships between edits and selection per minute, edit ratio, and duration of interruptions} %, and editions and selections.}
\label{tbl:p_value2}
\centering
\begin{tabular}{l | p{0.7cm} | p{1.9cm} | p{1.9cm} } 
   & none & short &  prolonged  \\  
  \hline
  \multicolumn{4}{c}{\textbf{Edits}} \\
  \hline
  mean & 6.96 &	2.98 & 1.76 \\ 
  % \cline{3-4} 
  %t.test & $\hookrightarrow$& \multicolumn{2}{c}{$<$ 2.2e-16} \\
   \cline{3-4} 
  U-test & $\hookrightarrow$ & \multicolumn{2}{c}{$<$ 2.2e-16}  \\

  \cline{3-4} 
  Cohen's $d$ & $\hookrightarrow$	& \textbf{1.23} & \textbf{2.11}   \\
  \hline

  \multicolumn{4}{c}{\textbf{Selections}} \\
  \hline 
  mean & 5.20 &	3.33 & 2.47 \\ 
  % \cline{3-4} 
  %t.test & $\hookrightarrow$& 2.089e-08 & 8.175e-14 \\
   \cline{3-4} 
  U-test & $\hookrightarrow$ & 6.197e-13 & $<$ 2.2e-16 \\

  \cline{3-4} 
  Cohen's $d$ & $\hookrightarrow$	& \textbf{0.63} & \textbf{1.24}  \\  
  \hline
  \multicolumn{4}{c}{\textbf{Edit ratio}} \\
  \hline 
  mean & 0.55 & 0.40 & 0.34\\ 
%   \cline{3-4} 
%  t.test & $\hookrightarrow$&  4.20E-12 & $<$ 2.2e-16 \\
   \cline{3-4} 
  U-test & $\hookrightarrow$ & 3.72e-14 & $<$ 2.2e-16  \\
  \cline{3-4} 
  Cohen's $d$ & $\hookrightarrow$ & \textbf{0.86} & \textbf{1.32}\\
\hline

%\begin{table}[ht!] \renewcommand{\arraystretch}{1.3} \caption{Effect size of the relationship between duration of interruptions and  ratio of edition.} \label{tbl:test_ratio2} \centering
%\begin{tabular}{l | p{0.7cm} | C{1.9cm} | C{1.9cm} } 
%   & none & short &  prolonged  \\  
%  \hline
%  \multicolumn{4}{c}{\textbf{Edits}} \\
%  \hline
%  mean & 0.55 & 0.40 & 0.34\\ 
%   \cline{3-4} 
%  t.test & $\hookrightarrow$&  4.20E-12 & $<$ 2.2e-16 \\
%   \cline{3-4} 
%  wilcox & $\hookrightarrow$ & 3.72E-14 & $<$ 2.2e-16  \\
%  \cline{3-4} 
%  effect size & $\hookrightarrow$ & \textit{0.86} & \textit{1.32}\\
%\end{tabular} \end{table}

Similar results occurs when we look at the edit ratio (Figure \ref{fig:box_dur_events}): the edit ratio is smaller in sessions that have prolonged interruptions, compared to the ones that only have short interruptions. At the bottom of Table \ref{tbl:p_value2}, we show the statistical and practical significance of these results. As before, we observe large effect sizes when comparing sessions who do not have interruptions with ones that do have, and larger effect sizes for sessions with at least one longer interruption.

These findings seem to indicate that the inverse relationship between productivity and time of duration is more pronounced in session with at least one longer interruption. This agrees with the literature for information workers.

On the other hand, before the interruption there is a preparation phase when the it is imminent or expected. On this phase the programmer might leave notes or comments in the code to recover the context after the interruption \cite{PD10}. Also, when the interruption happened because of a problem found by the programmer, usually he tries to solve it by reading the code, using the debugger or reading comments in the control version tool; after this resources are depleted, the next option is to ask to teammates or other external resources. The latest activities can represent an interruption of work.

In order to better understand the relationship between work fragmentation and productivity, we need to delve deeper and perform local analyses of the development sessions. In the first step, we wanted to summarize how the activity of users is distributed over time in sessions which have no interruptions, short interruptions, and prolonged interruptions.

To summarize the development sessions, we have resized each time series to a single $size$, using local means. We have used $size = 10$, that is, dividing each session in 10 chunks of equal time, since because it yielded better global visualization (better uniformity) of the user interaction along a session. For each chunk, we choose the median value of edits and selections for each group, and compose one summary time series for each group; these are the time series displayed in Figure \ref{fig:bins_int}.

These results support our previous observations at a finer level, and corroborate the literature saying that time is needed to start or get back on task.

\begin{figure}[!ht]
\centering
\begin{tabular}{c}
\minipage{0.33\textwidth}
\includegraphics[width=\linewidth,clip=, angle=0]{figures/without_int_med} 
\endminipage\hfill
\minipage{0.33\textwidth}
\includegraphics[width=\linewidth,clip=, angle=0]{figures/with_int1_med} 
\endminipage\hfill
\minipage{0.33\textwidth}
\includegraphics[width=\linewidth,clip=, angle=0]{figures/with_int2_med} 
\endminipage\hfill
\end{tabular}
\caption{Global representation of a session with: no interruptions (top); only short interruptions (middle); and at least one prolonged interruption (bottom). }
\label{fig:bins_int}
\end{figure}

Having presented the global effect of the interruptions over the user productivity, we now focus on the local activity before and after interruptions. We take a maximum real time interval of 30 minutes around each detected interruption, obtaining a set of 26988 time series subsequences. Then, we compute the median of all these subsequences as a generic local representation (Figure \ref{fig:all_ints}). We also plot with dashed lines the median values of edits and selections per minute in the sessions with interruptions in order to give more context to the observed values.
%where another interruption  should not have been found. 
Below we describe some observations:
\begin{itemize}
\item \textit{In the center,} we find the interruption point. There is clearly a negative effect on the time series, as the area before and after the interruption is the area with the lowest activity. The activity is well below the median activity of time series with interruptions, showing that the effect is indeed more pronounced near interruptions.

\item \textit{On the right,} the trend of the time series increases steadily. We hypothesize that the programmer is immersing again into the programming task, increasing progressively the activity as represented by the number of edits and selections. We see that in the average case, it reaches the median activity 12 to 18 minutes after the interruption. It then rises further than the session median, which is not surprising, as we expect higher than median activity further away from interruptions.

%We called \textit{recuperation time} the first intersect point between edition and navigation. In this sense, the  average recuperation time of an interruption occurs in the minute 12. 
\item \textit{On the left,} we observe that the number of events near the interruption also goes down well below the median value. This is might be because of two reasons: the programmer could have found a problem while coding, who at first will try to solve it by reading the code, switching out from the IDE, navigating the call stack and debugging, before going to ask another teammate, according to LaToza et al. \cite{LVD06}; this set of actions end up with an interruption and reduce the observed activity within the IDE. Also, when the interruption is imminent or expected, the programmer make use of different suspension strategies like writing physical notes, making a mental note or leaving a reminder cue on the code or window, as mentioned by Parnin and DeLine \cite{PD10}. These activities reduce the activity before the gap and they are seen in the interaction data mostly as selection events. The latest can be better visualized in Figure \ref{fig:all_ints_udc} with the UDC data as a significant decrease of edit events, being below the selection events approximately 10 minutes before the interruption.
%\item \textit{On the left,} we observe that the number of events near the interruption also goes down well below the median value. This is because work fragmentation is not only due to interruptions: programmers may for instance switch away from the IDE to look up information on the web or request it from other programmers. They are likely to do so because they are stuck, which would manifest itself by lower activity before the activity gap. We investigate this further below.

\begin{figure}[!ht]
\centering
\begin{tabular}{c}
\includegraphics[width=0.6\linewidth,clip=, angle=0]{figures/all_ints} 
\end{tabular}
\caption{Local effect of an interruption in the user activity.}
\label{fig:all_ints}
\end{figure}

\begin{table}[ht!]
\tiny
\renewcommand{\arraystretch}{1.3}
\caption{Effect size of the ratio of edits around the interruption.}
\label{tbl:ratio_all}
\centering
\begin{tabular}{l | p{0.7cm} | p{1.9cm} | p{1.9cm} } 
   & [1-5] & [6-10] & [11-15] \\  
  \hline
  \multicolumn{4}{c}{\textbf{before}} \\
  \hline
  mean & 0.34 &	0.37 &	0.39 \\ 
   \cline{3-4} 
  U-test & $\hookrightarrow$ & \multicolumn{2}{c}{$<$ 2.2e-16}  \\

\begin{table}[ht!]
\caption{Local Effect of Interruption. }
\label{tbl:local_effect}
\centering
\begin{tabular}{p{4cm} | c}
effect & pattern \\
\hline
\textit{negative} (45\%): when the frequency of edit events decreases after the interruption (Cohen's $d < -0.2$)
	& \includegraphics[valign=m,scale=0.3]{figures/neg_ints} \\
\textit{positive} (44\%): when the frequency of edit events increases after the interruption (Cohen's $d > 0.2$)
	& \includegraphics[valign=m,scale=0.3]{figures/pos_ints} \\
\textit{neutral} (11\%): when there is no well defined effect before or after the interruption ($abs(d) <= 0.2$)
	& \includegraphics[valign=m,scale=0.3]{figures/neu_ints} 
\end{tabular}
\end{table}

\begin{table}[ht!]
\tiny
\renewcommand{\arraystretch}{1.3}
\caption{Attributes in UDC data. }
\label{tbl:att_udc}
\centering
\begin{tabular}{|p{2.5cm}|p{7cm}|} 
  \hline 
attribute & description \\  
  \hline 
    \hline 
userId &  Unique number that identifies a user \\
what & The action of the event (deactivated, activated, opened, executed, etc.)  \\
kind & What kind of event was executed (workbench, view, command, etc.)  \\
bundleId & Description of the event's package  \\
bundleVersion & Version of the bundle's event  \\
description & Description of the event\\
datetime & Date and time in UNIX format\\
  \hline
\end{tabular}

\subsection{Selection and transformation}
We used the pre-processed data by Murphy-Hill, that is published on Google BigQuery \cite{SnipesETALASD}. This is an alternative version of the original UDC dataset, that is cleaned and preprocessed, so that the transformation phase is simple and focused on our needs. Due to the magnitude of the dataset we only worked on a fragment of it. We took a random sample of 1,000 users that is larger than the dataset we were able to obtain for the Mylyn study.

After obtaining the data, the next step was to add the attributes required by every event in order to match as closely as possible the format of the Mylyn data; this allowed us to reuse a part of our analysis. Hence, we added fields corresponding to the type of the event (edition or selection), its duration (time elapsing between one event and the next one, used to determine where interruptions take place and sessions end) and an ID to identify the different working sessions present on the data. For the latter, we sorted the data by \textit{userId} and \textit{datetime}. This was required because by default the user's data is mixed and we need it not only chronologically correct but also sorted by users to tag the working sessions of every user without interferences.

Once the events were classified and sorted, we split the dataset into sessions, defining a session as a period of registered work between interruptions of 8 hours or larger, as we did with Mylyn. We split the data by users to avoid a conflict between the last event of a user and the first one of another. After this, we obtained a total of 28,989 different sessions; but after a filter to keep only those with at least 30 minutes of productive time (as we did with the Mylyn dataset), we ended up with 15,825 useful sessions, in contrast with the 4,284 we had from Mylyn. Some statistics about the sessions can be seen in the Table \ref{tbl:stats_sessions_udc}, and a histogram of the duration of the sessions on UDC is shown in the Figure \ref{fig:hist_session_duration_udc}, which ignores outliers by setting a threshold of three times the standard deviation.

\begin{table}[ht!]
\tiny
\renewcommand{\arraystretch}{1.3}
\caption{Proportion of events. }
\label{tbl:prop_events_udc}
\centering
\begin{tabular}{|p{1.7cm}|p{2 cm}|} 
  \hline 
event & percentage \\  
  \hline 
    \hline 
edition &  $34.6\%$  \\
selection & $31.1\%$ \\
system & $31.5\%$  \\
control & $0.8\%$  \\
\hline
\end{tabular}
\end{table}

\begin{table}[ht!]
\tiny
\renewcommand{\arraystretch}{1.3}
\caption{Sample of classified events. }
\label{tbl:ej_events_udc}
\centering
\begin{tabular}{|p{5.5cm}| p{2 cm} |} 
  \hline 
description & type \\  
  \hline 
    \hline 
org.eclipse.ui.edit.delete & edition \\
org.eclipse.ui.file.save & edition \\
org.eclipse.ui.edit.undo & edition \\
org.eclipse.jdt.ui.CompilationUnitEditor & selection \\
org.eclipse.ui.console.ConsoleView & selection \\
org.eclipse.debug.ui.commands.StepOver & selection \\ 
org.eclipse.ui.workbench & system \\
org.eclipse.team.sync.views.SynchronizeView & control \\
\hline
\end{tabular}
\end{table}

After processing the data, we were able to reuse the same data analysis pipeline we used for Mylyn, in order to verify whether the results hold with this other dataset.

\section{Results of RQ1, RQ2, and RQ3 on UDC}

>>>>

\begin{table}[ht!]
\small
\renewcommand{\arraystretch}{1.3}
\caption{Kinds of interaction events~\cite{KM06} in Mylyn. }
\label{tbl:kind_event}
\centering
\begin{tabular}{|p{1.7cm}|p{1.3cm}|p{8cm}|} 
  \hline 
kind & mode & description \\  
  \hline 
    \hline 
selection & direct &  Editor and view selections via mouse \\
edit & direct & Textual and graphical edits  \\
command & direct & Operations such as saving, building, preference setting  \\
manipulation & direct & Direct manipulation of interest  \\
propagation & indirect & Interaction propagates to structurally related elements  \\
prediction & indirect & Capture of potential future interaction events \\
  \hline
\end{tabular}

@@

\begin{table}[ht!]
\small
\renewcommand{\arraystretch}{1.3}
\caption{ Example of interaction history. }
\label{tbl:sample_event}
\centering
\begin{tabular}{|c|l|c|c|c|} 
  \hline 
    & StartTime & EndTime & EventKind &  Method  \\
  \hline
  \hline
  1 & 10:30:00 & - & selection & m1  \\
  2 & 10:30:40 & - & manipulation & - \\
  3 & 10:30:40 & - & edit & m1  \\
  4 & 10:31:03.700 & - & selection & m1  \\
  5 & 10:31:03.800 & - & selection & m2  \\	  
  6 & 10:31:03.850 & - & selection & m3  \\	  
  7 & 10:32:05 & 10:33:07 & edit (5) & m2  \\	  
  8 & 10:33:10 & - & prediction & - \\
  \hline  
\end{tabular}
\end{table}

@@

\subsection{Selection}
We used the phases of the process \textit{Knowledge Discovery in Databases (KDD~\cite{FPP96})} to discover useful knowledge from our collection of data: we first select data, then preprocess it, before transforming it to time series.

Our purpose is the analysis of work fragmentation and interruptions in software development. Therefore, we focused only in the interaction data of the user with the program elements. The  variables of interest for our study are~\cite{RD13}:
\begin{enumerate}
	\item Duration: the time that a programmer took to perform a programming task. 
	\item Edition:  the amount of code changes that were necessary to perform said task. %It is associated to the  edit events.
	\item Navigation: how many program elements were consulted in a task; represented by selection events.
	\item Edit ratio: as in previous studies \cite{KM06}, the ratio of edits over edits and selections is an indicator of more efficient work since program exploration is reduced.
\end{enumerate}

@@

\begin{table}[hb!]
	\small
	\renewcommand{\arraystretch}{1.3}
	\caption{Number of events after preprocessing.}
	\label{tbl:mass_events}
	\centering
	\begin{tabular}{l | c | c | c } 
		type event & before & after & \% variation  \\  
		\hline 
		edit &	 452,236 & 475,554 &	+5\%   \\
		selection &	658,768 & 527,652 & -20\%   \\
		\hline
		total & 1,111,004 & 1,003,206 & -10\%  \\
	\end{tabular}
\end{table}

@@

\begin{table}[ht!]
	\small
	\renewcommand{\arraystretch}{1.3}
	\caption{Prevalence of interruptions according to their duration time. }
	\label{tbl:by_duration}
	\centering
	\begin{tabular}{p{3cm}|p{0.8cm}|p{6cm}} 
		duration & ratio & examples  \\
		\hline   
		$[3  - 12 min \rangle$ & 69\% & short pause, answer a question, thinking, looking for interruption  \\ 
		\hline 
		$[12  - 30 min \rangle$ & 18\% & coffee break, short meeting, extended interruption  \\
		$[30 min - 2 hr \rangle$  & 9\% & lunch break, a meeting  \\
		$[2 hr - 8 hr \rangle$ & 4\% & extended meeting \\
		%$[8 hr - \infty \rangle$ & 0.14 & \textit{a long interruption}   \\
		%$[8 hr -  \infty \rangle$ & 0.06  &  a weekend \\
		%$[2 day - \infty \rangle$ & 0.04 & a long holiday \\ 
	\end{tabular}
\end{table}

\begin{figure}[t]
	\centering
	\begin{tabular}{c}
		\includegraphics[width=.6\linewidth,clip=, angle=0]{figures/trans} 
	\end{tabular}
	\caption{Example of how to compress a time series. }
	\label{fig:trans}
\end{figure}

\begin{table}[ht!]
	\small
	\renewcommand{\arraystretch}{1.3}
	\caption{Attributes in UDC data. }
	\label{tbl:att_udc}
	\centering
	\begin{tabular}{|p{2.5cm}|p{10cm}|} 
		\hline 
		attribute & description \\  
		\hline 
		\hline 
		userId &  Unique number that identifies a user \\
		what & The action of the event (deactivated, activated, opened, executed, etc.)  \\
		kind & What kind of event was executed (workbench, view, command, etc.)  \\
		bundleId & Description of the event's package  \\
		bundleVersion & Version of the bundle's event  \\
		description & Description of the event\\
		datetime & Date and time in UNIX format\\
		\hline
	\end{tabular}

\subsection{Selection}
We used the pre-processed version of the data that is published on Google BigQuery, by Murphy-Hill \cite{SnipesETALASD}. This is an alternative version of the original UDC dataset, that is cleaned and preprocessed, so that the transformation phase is simple and focused on our needs. Due to the magnitude of the dataset we only worked on a fragment of it. We took a random sample of 1,000 users that is larger than the dataset we were able to obtain for the Mylyn study.

\subsection{Preprocessing}
After obtaining the data, the next step was to add the attributes required by every event in order to match as closely as possible the format of the Mylyn data; this allowed us to reuse a part of our analysis. Hence, we added fields corresponding to the type of the event (edition or selection), its duration (time elapsing between one event and the next one, used to determine where interruptions take place and sessions end) and an ID to identify the different working sessions present on the data. For the latter, we sorted the data by \textit{userId} and \textit{datetime}. This was required because by default the user's data is mixed and we need it not only chronologically correct but also sorted by users to tag the working sessions of every user without interferences.

It was also required to fill the description for the events that indicate the activation or deactivation of the workbench of Eclipse, but as they were the only events with this error this task was trivial. However, it was needed to classify the events, which was a complicated task.

\begin{table}[ht!]
	\small
	\renewcommand{\arraystretch}{1.3}
	\caption{Proportion of events. }
	\label{tbl:prop_events_udc}
	\centering
	\begin{tabular}{|p{1.7cm}|p{2 cm}|} 
		\hline 
		event & percentage \\  
		\hline 
		\hline 
		edition &  $34.6\%$  \\
		selection & $31.1\%$ \\
		system & $31.5\%$  \\
		control & $0.8\%$  \\
		\hline
	\end{tabular}
\end{table}

\begin{table}[ht!]
	\small
	\renewcommand{\arraystretch}{1.3}
	\caption{Sample of classified events in UDC. }
	\label{tbl:ej_events_udc}
	\centering
	\begin{tabular}{|p{6cm}| p{2 cm} |} 
		\hline 
		description & type \\  
		\hline 
		\hline 
		org.eclipse.ui.edit.delete & edition \\
		org.eclipse.ui.file.save & edition \\
		org.eclipse.ui.edit.undo & edition \\
		org.eclipse.jdt.ui.CompilationUnitEditor & selection \\
		org.eclipse.ui.console.ConsoleView & selection \\
		org.eclipse.debug.ui.commands.StepOver & selection \\ 
		org.eclipse.ui.workbench & system \\
		org.eclipse.team.sync.views.SynchronizeView & control \\
		\hline
	\end{tabular}
\end{table}

\subsection{Transformation}
Once the events were classified and sorted, we split the dataset into sessions, defining a session as a period of registered work between interruptions of 8 hours or larger, as we did with Mylyn. We split the data by users to avoid a conflict between the last event of a user and the first one of another. After this, we obtained a total of 28,989 different sessions; but after a filter to keep only those with at least 30 minutes of productive time (as we did with the Mylyn dataset), we ended up with 15,825 useful sessions, in contrast with the 4,284 we had from Mylyn. At this point, we had the same format of sessions with both datasets. Some statistics about the sessions can be seen in the Table \ref{tbl:stats_sessions_udc}, and a histogram of the duration of the sessions on UDC is shown in the Figure \ref{fig:hist_session_duration_udc}, which ignores outliers by setting a threshold of three times the standard deviation.

%After processing the data, we were able to reuse the same data analysis pipeline we used for Mylyn, in order to verify whether the results hold with this other dataset.

%\section{Preview Phases for Knowledge Discovery}
%We used the phases of the process \textit{Knowledge Discovery in Databases (KDD~\cite{FPP96})} to discover useful knowledge from our collection of data: we first select data, then preprocess it, before transforming it to time series. We close this section by presenting the metrics we use in this study. %  (see Figure~\ref{fig:esq}). 
%

\section{RQ1: Relationship between Interruptions and Productivity}

\subsection{Metrics Used in This Study}
Before introducing the results about the first research question, it is convenient to present the metrics used in this study.
We use the following six metrics to measure interruptions and productivity, while controlling for the unproductive time in a session, the length of time of the session itself, and the efficiency of the development that took place during the session:

@@

%\subsection{Effect in the productive work time}

@@

%First, we analyzed the number of interruptions over the productivity. In this sense, 
We split the data in five groups: The first group contain all the sessions without interruptions (\textit{none}). For the others groups, we have considered four ranges of number of interruptions delimited by their quartiles; the results differ between Mylyn and UDC, existing more interruptions in the former (Table \ref{tbl:quartil_int}). Then, for each group, we display the distribution of events per minute and edit ratio via boxplots. We can see the results with Mylyn in Figure \ref{fig:box_int_events}), and for UDC in Figure \ref{fig:box_int_events_udc}.

In the results with Mylyn, we observe a large difference between the sessions without interruptions and the ones who do. Further, we observe that the rate of events per minute decreases slightly when the session has more interruptions. Therefore, we can intuit that the relationship between number of interruptions and the productivity indicators that are edits and selections, tends to be inversely proportional.

With UDC, the results show a similar pattern in comparison with the results with Mylyn. When facing zero interruptions the editions and selections per minute are greater than with one or more interruptions, and as more interruptions occur this metrics gradually decrease. The difference between sessions with no interruptions and at least one is less marked in the edit per minute metric, but the selections per minute shows the same effect as with Mylyn.

\begin{table}[ht!]
\small
\renewcommand{\arraystretch}{1.3}
\caption{Thresholds used to group sessions based on their number of interruptions} %separate the data in four percentages groups regarding the number of interruptions in each trace.}
\label{tbl:quartil_int}
\centering
\begin{tabular}{l | p{0.6cm} | p{0.6cm} | p{0.6cm}} 
     & $25\%$ & $50\%$ & $75\%$ \\  
  \hline 
  Mylyn &  3 & 5 & 10  \\ 
  \hline  
  UDC &  2 & 4 & 7  \\  
\end{tabular}
\end{table}

\begin{figure}[!ht]
\centering
\begin{tabular}{c}
\minipage{0.33\textwidth}
\includegraphics[width=\linewidth,clip=, angle=0]{figures/box_int_events_edit}  
\endminipage\hfill
\minipage{0.33\textwidth}
\includegraphics[width=\linewidth,clip=, angle=0]{figures/box_int_events_sel}  
\endminipage\hfill
\minipage{0.33\textwidth}
\includegraphics[width=\linewidth,clip=, angle=0]{figures/box_int_ratio} 
\endminipage\hfill
\end{tabular}
\caption{Boxplots showing the relation between the number of edits and selection events per minute, the edit ratio, and the number of interruptions in Mylyn. }
\label{fig:box_int_events}
\end{figure}

\begin{figure}[!ht]
	\centering
	\begin{tabular}{c}
		\minipage{0.33\textwidth}
		\includegraphics[width=\linewidth,clip=, angle=0]{figures/udc_box_int_events_edit}  \\
		\endminipage\hfill
		\minipage{0.33\textwidth}
		\includegraphics[width=\linewidth,clip=, angle=0]{figures/udc_box_int_events_sel}  \\
		\endminipage\hfil
		\minipage{0.33\textwidth}
		\includegraphics[width=\linewidth,clip=, angle=0]{figures/udc_box_int_events_ratio} \\
		\endminipage\hfill
	\end{tabular}
	\caption{Boxplots showing the relation between the number of editions and selections per minute, the edit ratio, and the number of interruptions in UDC. }
	\label{fig:box_int_events_udc}
\end{figure}

Beyond visual inspection, we also quantify the statistical and the practical significance of these observations. First, in the results with Mylyn, all the differences observed are statistically significant with very low p-values (see Table~\ref{tbl:p_value}) according to the Mann-Whitney U-test. This is not surprising, given the shape of the boxplots and the size of the samples.

@@

On the other hand, with the results in UDC, an important difference is that the effect overall appears to be less pronounced in UDC and we can see it on the effect size tests shown in the Table \ref{tbl:p_value_udc}. Most of the observed effects are medium, while in the Mylyn dataset the effects were large. Note that despite the differences in effect sizes, the differences are still significant and it is possible to reach the same conclusions.

@@

\begin{table}[ht!]
\tiny
\renewcommand{\arraystretch}{1.3}
\caption{Effect size and significance of the relationship between number of edit per minute, selection per minute, edit ratio, and number of interruptions on Mylyn data} 
\label{tbl:p_value}
\centering
\begin{tabular}{l | p{0.75cm} | p{1.2cm} | p{1.2cm} | p{1.2cm} |p{1.2cm}}

@@
